---
title: "pedPMS Proteomics 1: Import, QC and reference-channel IRS (workshop)"
author: "Workshop"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    toc-title: "Table of contents"
    code-fold: true
    code-summary: "Code"
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    theme: cosmo
execute:
  warning: false
  message: false
  cache: false
editor: source
---

<!--
  WORKSHOP EDITION (conflict-hardened). This is the first of two reports. It reads
  the raw workshop files, builds the sample metadata, checks data quality, runs
  internal reference scaling (IRS) using the pooled reference channel, and saves
  one tidy object for the second report to analyse.

  Every code chunk starts with "# AIM:" saying WHY it exists, and the harder
  functions carry a one-line plain-language comment. Read a pipe (%>%) as the
  word "then": x %>% f() %>% g() means take x, then do f, then do g.

  WHICH INPUT FILE. Set params_local$data_txt to one of:
    - pedPMS_data_preIRS.txt  : reference present, plexes NOT yet aligned. IRS
                                below does real work. Use this to LEARN IRS.
    - pedPMS_data_postIRS.txt : reference present, already aligned. IRS below is
                                almost a no-op. Use this to VERIFY IRS was done.
    - pedPMS_data.txt         : no reference channel. IRS cannot be recomputed;
                                the script verifies alignment diagnostically.
  The same code runs on any of them; only the story changes.
-->

```{r setup}
# AIM: load packages, resolve namespace conflicts, set the reproducibility seed,
# and define paths, thresholds and small helpers used later.

library(tidyverse)             # dplyr / ggplot2 / stringr / tidyr / purrr
library(readxl)                # read_excel(): read .xlsx spreadsheets
library(matrixStats)           # fast rowSds / rowVars / colSums on matrices
library(cowplot)               # cleaner ggplot themes
library(ggExtra)               # ggMarginal(): marginal histograms on a scatter
library(ggpubr)                # stat_cor(): correlation printed on a plot
library(ggrepel)               # non-overlapping text labels
library(rstatix)               # tidy statistical tests
library(pcaMethods)            # pca(): PCA that tolerates missing values
library(DT)                    # datatable(): interactive HTML tables
library(SummarizedExperiment)  # container bundling matrix + metadata
library(here)                  # here(): locate the project root portably

# --- Namespace conflicts -------------------------------------------------------
# Bioconductor packages (here SummarizedExperiment, which attaches S4Vectors,
# IRanges, BiocGenerics, MatrixGenerics ...) export functions with the same names
# as the tidyverse and base R: select, filter, slice, rename, rowSds, rowVars,
# table, and more. The conflicted package turns each such clash into an explicit
# error the first time you CALL the ambiguous name, so a script that runs today
# can fail tomorrow when a new package is attached. Resolve it once, by policy:
#   1) dplyr wins every conflict it is part of (all the tidyverse verbs).
#   2) pin the base / matrixStats functions we call by bare name that a
#      Bioconductor generic would otherwise mask.
# If you attach another package later and hit a "conflict" error, add one line:
#   conflicts_prefer(thePackageYouWant::theFunction)
library(conflicted)            # needs conflicted >= 1.1.0 for conflict_prefer_all()
suppressMessages({
  conflict_prefer_all("dplyr", quiet = TRUE)
  conflicts_prefer(
    matrixStats::rowSds, matrixStats::rowVars,      # vs MatrixGenerics
    base::table, base::which.max, base::which.min,  # vs BiocGenerics
    base::rowMeans, base::colMeans,
    base::rowSums,  base::colSums,
    stats::median,  stats::sd
  )
})

set.seed(20240101)   # fixes randomness so results are identical each run

# --- Project location ----------------------------------------------------------
# Base folder for the whole workshop, so data_raw/ and results/ always resolve to
# the same place whether you Render the .qmd, run it line by line, or clone the
# repo onto a different machine. here::here() finds the project root by looking
# for a .git folder or an .Rproj file above the current working directory, so
# this line needs no editing after `git clone` on Windows, Mac or Linux. If you
# ever run chunks from a subfolder, keep the working directory at the repo root
# (open the .Rproj, or in RStudio: Session > Set Working Directory > To Project
# Directory) so here() has something to find.
base_dir <- here::here()

# All paths and thresholds in one place. file.path() joins with "/" for you.
# The metadata is now read from CSV (pedPMS_Samples.csv and
# pedPMS_Protein_samples_Batch_1.csv). If you prefer the .xlsx sheets, point these
# back at the .xlsx files and swap the two read_csv() calls in metadata-import
# back to read_excel().
params_local <- list(
  data_txt      = file.path(base_dir, "data_raw", "pedPMS_data_preIRS.txt"), # pre/post/plain
  sample_sheet  = file.path(base_dir, "data_raw", "pedPMS_Samples.csv"),
  prep_sheet    = file.path(base_dir, "data_raw", "pedPMS_Protein_samples_Batch_1.csv"),
  out_dir       = file.path(base_dir, "results"),
  max_missing   = 50,   # drop a protein if missing in more than this % of samples
  ref_channel   = 11,   # channel carrying the pooled reference in every plex
  tmt_tags      = c("126", "127N", "127C", "128N", "128C",
                    "129N", "129C", "130N", "130C", "131")
)

# AIM: build the output folder tree. Everything the report writes goes here:
# the saved object at the top level, CSV tables under results/tables, and image
# files under results/figures. recursive = TRUE makes parent folders as needed.
dirs <- list(
  root    = params_local$out_dir,
  tables  = file.path(params_local$out_dir, "tables"),
  figures = file.path(params_local$out_dir, "figures"))
for (p in dirs) dir.create(p, showWarnings = FALSE, recursive = TRUE)

# reused interactive-table helper.
dt_table <- function(x) {
  DT::datatable(x, filter = "top",
                options = list(autoWidth = FALSE, scrollX = TRUE),
                class = "compact hover row-border stripe dt-left cell-border nowrap")
}

# save a data frame to results/tables as CSV, and return it unchanged so it can
# still be piped into dt_table() for display. One call both saves and shows.
save_table <- function(x, name) {
  readr::write_csv(x, file.path(dirs$tables, name))
  x
}

# save a ggplot to results/figures as PNG, and return it unchanged so it still
# displays in the report.
save_ggplot <- function(p, name, w = 8, h = 5) {
  ggplot2::ggsave(file.path(dirs$figures, name), p, width = w, height = h,
                  dpi = 150, bg = "white")
  p
}
```

# Introduction

This report imports the pedPMS TMT proteomics dataset, checks its quality, and
normalises it across plexes using a pooled reference channel. The dataset is
synthetic and built for teaching, but its structure matches a real TMT
experiment: 66 biological samples labelled with TMT tags, run in 7 plexes, each
plex injected twice into the mass spectrometer, plus one pooled reference channel
per plex used to align the plexes.

Two properties of the data shape everything below and are worth stating now.

Samples were assigned to plexes in tube order, so disease subtype tracks plex.
That confounding is quantified in its own subsection and it limits how subtype
contrasts can be read later.

Six of the 66 samples are healthy controls, and all six sit in plex 7 together
with three disease samples. Plex 7 is therefore the one place a healthy-versus-
disease comparison carries no plex effect, which the second report uses.

# Data Carpentry and Metadata

```{r metadata-import}
# AIM: turn the two CSVs into one clean metadata table, meta.aml, with one row per
# biological sample. Robust version: reads CSV with every column pinned to text
# (so tube ids like 1, 2 are not re-guessed as numbers), detects the sample rows
# rather than assuming exactly 66, trims ids, and names any mismatch instead of a
# bare stopifnot.

# normalise an id: trim whitespace, strip a trailing .0 / .00, empty -> NA.
norm_id <- function(x) {
  x <- trimws(as.character(x))
  x <- sub("\\.0+$", "", x)
  x[x == "" | x == "NA"] <- NA_character_
  x
}
all_text <- readr::cols(.default = readr::col_character())   # pin every column to text

# labelling layout (Listen sheet, exported to CSV): no header row, so col_names =
# FALSE. Rows 1-3 are reagent notes; drop them, keep the columns we use, fill the
# plex label down its block, then keep only rows that carry a tube id.
sheet_raw <- readr::read_csv(params_local$sample_sheet, col_names = FALSE,
                             col_types = all_text, show_col_types = FALSE)

sample_layout <- sheet_raw[-(1:3), c(1, 3, 4, 5, 6)]
colnames(sample_layout) <- c("label_id", "tmt_tag", "plex_label",
                             "plex_id", "tube_id")
sample_layout <- sample_layout %>%
  fill(plex_label, plex_id) %>%                              # plex written once; copy down
  dplyr::filter(!is.na(tube_id) & trimws(tube_id) != "") %>% # real sample rows only
  dplyr::mutate(
    plex     = as.integer(str_remove(plex_label, "\\.Plex")),  # "7.Plex" -> 7
    plex_key = str_remove(plex_id, "^PX-"),                    # "PX-598" -> "598"
    channel  = match(tmt_tag, params_local$tmt_tags),          # tag -> position 1-10
    tube_id  = norm_id(tube_id))

# prep/sorting record (LIST sheet, exported to CSV): has a header row (col_names
# default TRUE). Keep rows that carry a tube id.
prep_meta <- readr::read_csv(params_local$prep_sheet,
                             col_types = all_text, show_col_types = FALSE) %>%
  dplyr::select(tube_id = 1, cell_number = 2, lysis_volume = 3, subtype = 4) %>%
  dplyr::filter(!is.na(tube_id) & trimws(tube_id) != "") %>%
  dplyr::mutate(cell_number = as.numeric(cell_number),
                tube_id     = norm_id(tube_id))

# safety gate: name the mismatched tubes instead of a bare stopifnot, so a broken
# join tells you exactly which samples are out of sync rather than failing blind.
only_layout <- setdiff(sample_layout$tube_id, prep_meta$tube_id)
only_prep   <- setdiff(prep_meta$tube_id, sample_layout$tube_id)
if (length(only_layout) || length(only_prep)) {
  stop("Tube ids differ between the two metadata files.\n",
       "  layout n = ", nrow(sample_layout),
       ", prep n = ", nrow(prep_meta), "\n",
       "  in layout only: ", paste(only_layout, collapse = ", "), "\n",
       "  in prep only:   ", paste(only_prep,   collapse = ", "), "\n",
       "  Use the matching versions of both files in data_raw/.",
       call. = FALSE)
}
stopifnot(!any(is.na(sample_layout$channel)),
          n_distinct(sample_layout$label_id) == nrow(sample_layout))

meta.aml <- sample_layout %>%
  dplyr::left_join(prep_meta, by = "tube_id") %>%
  dplyr::mutate(subtype = factor(subtype),
                group   = if_else(subtype == "control", "control", "disease")) %>%
  dplyr::select(label_id, tube_id, plex, plex_id, plex_key, tmt_tag, channel,
                subtype, group, cell_number, lysis_volume)
```

## Metadata summary

```{r metadata-summary}
# AIM: show the finished metadata table and two overview plots.
meta.aml %>% dt_table()
```

```{r metadata-plots, fig.height=4}
meta.aml %>%
  ggplot(aes(x = cell_number)) +
  geom_histogram(bins = 20) +
  cowplot::theme_minimal_grid() + cowplot::panel_border() +
  labs(title = "Viable cells at lysis", x = "cells (x10^6)", y = "")

meta.aml %>%
  dplyr::count(subtype) %>%
  ggplot(aes(x = fct_reorder(subtype, n), y = n, fill = subtype == "control")) +
  geom_col(show.legend = FALSE) + coord_flip() +
  scale_fill_manual(values = c("grey40", "steelblue")) +
  cowplot::theme_cowplot() +
  labs(x = "", y = "samples", title = "Subtype (controls highlighted)")
```

## Proteomics data

```{r proteomics-import}
# AIM: read the intensity matrix and split it into three parts: the biological
# matrix (prot.aml), the pooled-reference matrix (prot.ref), and the per-protein
# annotation. The reference is a technical channel, so it must NOT go through the
# sample metadata join (it has no patient behind it).

raw <- read_tsv(params_local$data_txt, skip = 1,
                locale = locale(encoding = "UTF-16LE"),
                show_col_types = FALSE)

# every "channel Rrep_plex" column
all_cols <- grep("^\\d+ R[12]_\\d+$", colnames(raw), value = TRUE)
# reference = the reference channel (11); biology = channels 1-10.
ref_pat  <- paste0("^", params_local$ref_channel, " R[12]_\\d+$")
ref_cols <- all_cols[str_detect(all_cols, ref_pat)]
sample_cols <- setdiff(all_cols, ref_cols)

prot.annotation <- raw %>%
  dplyr::select(uniprotid = `Uniprot IDs`, gene = `Gene names`,
                protein = `Protein names`, peptides = Peptides,
                unique_peptides = `Unique peptides`,
                coverage = `Sequence coverage [%]`,
                mol_weight = `Mol. weight [kDa]`, qvalue = `Q-value`, score = Score)

# biological matrix: proteins x 132 columns (66 samples x 2 injections)
prot.aml <- raw %>% dplyr::select(all_of(sample_cols)) %>% as.matrix()
rownames(prot.aml) <- raw$`Uniprot IDs`

# reference matrix: proteins x 14 columns (7 plexes x 2 injections)
prot.ref <- raw %>% dplyr::select(all_of(ref_cols)) %>% as.matrix()
rownames(prot.ref) <- raw$`Uniprot IDs`

# describe the reference columns so we can group them by plex during IRS.
ref.meta <- tibble(sample = ref_cols) %>%
  dplyr::mutate(replicate = str_extract(sample, "R[12]"),
                plex_key  = str_extract(sample, "\\d+$"))

# describe the 132 biological columns and attach sample info.
meta.cols <- tibble(sample = sample_cols) %>%
  dplyr::mutate(channel  = as.integer(str_extract(sample, "^\\d+")),
                replicate = factor(str_extract(sample, "R[12]"), levels = c("R1","R2")),
                plex_key = str_extract(sample, "\\d+$")) %>%
  dplyr::left_join(dplyr::select(meta.aml, label_id, plex_key, channel, subtype,
                                 plex, tube_id, cell_number),
                   by = c("plex_key", "channel"))

# checks: every biological column maps to a sample; no duplicate proteins; values
# positive and finite; every expected column present.
stopifnot(!any(is.na(meta.cols$label_id)),
          !any(duplicated(rownames(prot.aml))),
          all(prot.aml > 0, na.rm = TRUE),
          !any(is.infinite(prot.aml)),
          all(c(paste0(meta.aml$channel, " R1_", meta.aml$plex_key),
                paste0(meta.aml$channel, " R2_", meta.aml$plex_key)) %in%
              colnames(prot.aml)))
```

`r nrow(prot.aml)` protein groups across `r ncol(prot.aml)` biological columns
(66 samples, each injected twice) plus `r ncol(prot.ref)` reference columns
(7 plexes, each injected twice). Values are strictly positive linear-scale
reporter intensities, so a single log2 transform is used later and no pseudocount.

### Normalization

TMT normalisation has two steps. Sample loading (SL) equalises the total signal
per sample. Internal reference scaling (IRS) aligns the plexes to each other using
the pooled reference channel. We check SL and then run IRS.

#### Sample loading (verification)

```{r sl-check}
# AIM: check that each sample carries a similar total signal. If loading is
# equalised, the column sums are all close and their CV is small.

sl_plot <- tibble(sample = colnames(prot.aml),
       col_sum = colSums(prot.aml, na.rm = TRUE)) %>%
  dplyr::left_join(meta.cols, by = "sample") %>%
  ggplot(aes(x = fct_reorder(sample, plex), y = col_sum, fill = factor(plex))) +
  geom_col() + scale_fill_viridis_d(name = "Plex") +
  cowplot::theme_minimal_grid() + cowplot::panel_border() +
  labs(x = "", y = "summed reporter intensity") +
  theme(axis.text.x = element_blank())
save_ggplot(sl_plot, "00_sample_loading.png", w = 8, h = 4)
sl_plot

sl_cv <- sd(colSums(prot.aml, na.rm = TRUE)) / mean(colSums(prot.aml, na.rm = TRUE))
```

Column-sum CV is `r round(100 * sl_cv, 2)` %. Loading is already equalised across
all columns, so no sample-loading correction is applied here.

<!--
  IF SAMPLE LOADING WERE *NOT* EQUALISED (a large CV), you would add an SL step
  here, BEFORE IRS, scaling each column to a common target. Uncomment to use:

  target     <- exp(mean(log(colSums(prot.aml, na.rm = TRUE))))  # geometric-mean target
  sl_factors <- target / colSums(prot.aml, na.rm = TRUE)
  prot.aml   <- sweep(prot.aml, 2, sl_factors, "*")              # scale each column

  SL equalises totals within a plex; IRS (below) aligns plexes to each other.
  Doing them out of order mixes the two corrections.
-->

#### Internal reference scaling (IRS)

The pooled reference is the same sample in every plex, so any difference in its
measured value between plexes is pure batch effect. The ratio that pulls the
reference to a common level is the ratio that corrects the real samples in that
plex.

This section adapts to the input. If a reference channel is present (the pre-IRS
and post-IRS files), it computes IRS and applies it. If no reference channel is
present (the plain file), it cannot recompute IRS, so it instead verifies
alignment diagnostically, by checking that the spread of a protein between plexes
is no larger than the spread between the two injections of one plex.

```{r irs-reference-channel, fig.height=4}
# AIM: align the plexes. Two paths depending on the input:
#   with a reference channel -> compute and apply reference-channel IRS
#   without one              -> verify alignment diagnostically (no change made)
# has_reference is TRUE when channel-11 columns were found during import.
has_reference <- ncol(prot.ref) > 0

# helper: CV across plexes per protein (used by both paths).
cv_across_plexes <- function(m) {
  lin <- 2^m
  matrixStats::rowSds(lin, na.rm = TRUE) / rowMeans(lin, na.rm = TRUE)
}

if (has_reference) {

  ref.log <- log2(prot.ref)            # log2: multiplicative -> additive
  plex_keys <- unique(ref.meta$plex_key)

  # per-plex reference profile: average the two injections of each plex.
  ref.by.plex <- purrr::map(set_names(plex_keys), function(p) {
    cols <- ref.meta$sample[ref.meta$plex_key == p]
    rowMeans(ref.log[, cols, drop = FALSE], na.rm = TRUE)
  })
  ref.mat    <- do.call(cbind, ref.by.plex)          # proteins x plexes
  target.log <- rowMeans(ref.mat, na.rm = TRUE)       # per-protein target
  # per-plex, per-protein scaling in log2 = target - that plex's reference
  scale.log  <- purrr::map(set_names(plex_keys), function(p) target.log - ref.by.plex[[p]])
  ref.corrected <- do.call(cbind, purrr::map(plex_keys, ~ ref.by.plex[[.x]] + scale.log[[.x]]))

  irs_plot <- tibble(stage = rep(c("before IRS", "after IRS"), each = nrow(ref.mat)),
                     cv = c(cv_across_plexes(ref.mat), cv_across_plexes(ref.corrected))) %>%
    tidyr::drop_na() %>%
    mutate(stage = factor(stage, levels = c("before IRS", "after IRS"))) %>%
    ggplot(aes(x = stage, y = cv, fill = stage)) +
    geom_violin() + scale_y_continuous(labels = scales::percent) +
    cowplot::theme_cowplot() + theme(legend.position = "none") +
    labs(x = "", y = "reference CV across plexes",
         title = "Reference alignment before and after IRS")

  # apply the scaling to the biological matrix, per plex.
  prot.aml.log <- log2(prot.aml)
  for (p in plex_keys) {
    cols <- colnames(prot.aml.log)[str_detect(colnames(prot.aml.log), paste0("_", p, "$"))]
    prot.aml.log[, cols] <- prot.aml.log[, cols] + scale.log[[p]]
  }
  prot.aml <- 2^prot.aml.log          # IRS-corrected biology, carried forward
  median_abs_scale <- median(abs(unlist(scale.log)), na.rm = TRUE)
  irs_status <- paste0("Reference channel found. IRS applied; median absolute ",
                       "scaling ", round(median_abs_scale, 3), " log2 units.")

} else {

  # DIAGNOSTIC PATH: no reference channel, so IRS cannot be recomputed. Compare
  # the between-plex spread of each protein to the between-injection spread. If
  # the plexes are already aligned, the two are of similar size.
  block_means <- meta.cols %>%
    group_by(plex_key, replicate) %>%
    group_map(~ tibble(uniprotid = rownames(prot.aml),
                       block = paste(.y$replicate, .y$plex_key, sep = "_"),
                       mean_log2 = rowMeans(log2(prot.aml[, .x$sample, drop = FALSE])))) %>%
    bind_rows() %>% pivot_wider(names_from = block, values_from = mean_log2) %>% drop_na()
  between_block_sd <- block_means %>% dplyr::select(-uniprotid) %>% as.matrix() %>% matrixStats::rowSds()
  within_plex_sd <- map_dbl(unique(meta.cols$plex_key), function(p)
    sd(block_means[[paste0("R1_", p)]] - block_means[[paste0("R2_", p)]]))

  irs_plot <- tibble(source = c(rep("between plex blocks", length(between_block_sd)),
                                rep("between injections", length(within_plex_sd))),
                     sd = c(between_block_sd, within_plex_sd)) %>%
    ggplot(aes(x = source, y = sd, fill = source)) +
    geom_violin() + cowplot::theme_cowplot() + theme(legend.position = "none") +
    labs(x = "", y = "SD of per-protein mean log2",
         title = "Plex alignment (diagnostic, no reference channel)")
  irs_status <- paste0("No reference channel in this file. IRS was verified ",
                       "diagnostically and the matrix was left unchanged.")
}

save_ggplot(irs_plot, "01_irs_alignment.png", w = 6, h = 4)
irs_plot
```

`r irs_status` On the pre-IRS file the correction is clearly non-zero and the
before/after violins separate. On the post-IRS or plain file it is near zero or
absent, which is how the report shows alignment was already in place. A reference
channel can only correct a plex where the protein was measured, so proteins absent
from a whole plex stay uncorrected there; this is why missingness and IRS are
discussed together.

### Investigation

```{r log-and-merge}
# AIM: put data on log2 scale and average the two injections of each sample into
# one column, so prot.aml.merge has one column per biological sample (66).

prot.aml.log <- log2(prot.aml)

prot.aml.merge <- meta.aml %>%
  rowwise() %>%
  group_map(~ rowMeans(prot.aml.log[, c(paste0(.x$channel, " R1_", .x$plex_key),
                                        paste0(.x$channel, " R2_", .x$plex_key))],
                       na.rm = TRUE)) %>%
  bind_cols() %>% as.matrix()
colnames(prot.aml.merge) <- meta.aml$label_id
rownames(prot.aml.merge) <- rownames(prot.aml.log)
prot.aml.merge[is.nan(prot.aml.merge)] <- NA_real_   # 0/0 average -> proper NA
```

#### Missing values

```{r missing-per-protein}
# AIM: measure per-protein missingness and see how many exceed the 50% cutoff.
missingness_by_protein <- tibble(
  uniprotid = rownames(prot.aml.merge),
  no_missing = 100 * rowSums(is.na(prot.aml.merge)) / ncol(prot.aml.merge))

missingness_by_protein %>%
  ggplot(aes(x = no_missing, fill = no_missing > params_local$max_missing)) +
  geom_histogram(position = position_identity(), bins = 30) +
  cowplot::theme_minimal_grid() + cowplot::panel_border() +
  theme(legend.position = "none") +
  labs(x = "Percent missing", y = "",
       caption = paste0("blue = missing in more than ", params_local$max_missing,
                        "% of samples"))
```

```{r filter-impact}
# AIM: show how many proteins survive at different cutoffs, then apply the 50% one.
tibble(max_missing = c(50, 30, 20, 10, 0)) %>%
  dplyr::mutate(n_proteins = map_int(max_missing,
                              ~ sum(missingness_by_protein$no_missing <= .x))) %>%
  dt_table()

high_missingness_proteins <- missingness_by_protein %>%
  dplyr::filter(no_missing > params_local$max_missing) %>% pull(uniprotid)

prot.aml.merge.filtered <-
  prot.aml.merge[!rownames(prot.aml.merge) %in% high_missingness_proteins, ]
```

#### Replicate agreement

```{r replicate-agreement}
# AIM: measure how well a sample's two injections agree. High agreement confirms
# they are technical repeats, so averaging them is valid.
replicate_agreement <- meta.aml %>%
  rowwise() %>%
  dplyr::mutate(
    r = cor(prot.aml.log[, paste0(channel, " R1_", plex_key)],
            prot.aml.log[, paste0(channel, " R2_", plex_key)], use = "complete.obs")) %>%
  ungroup() %>%
  dplyr::select(label_id, tube_id, subtype, plex, r)

# save the per-sample QC table (replicate agreement) to results/tables.
replicate_agreement %>%
  dplyr::mutate(r = round(r, 4)) %>%
  save_table("sample_qc_replicate_agreement.csv") %>%
  dt_table()
```

#### Pattern of missingness

```{r missingness-blockwise}
# AIM: describe the missingness pattern per plex-run block. Most missingness is
# blockwise (a protein is measured in a whole plex block or in none of it), which
# is why we filter rather than impute. Note the "partly_missing" column: it is
# small but not zero, so the "whole plex" description is a good approximation, not
# an absolute rule.
meta.cols %>%
  group_by(plex_key, replicate) %>%
  group_modify(function(g, key) {
    m <- is.na(log2(prot.aml)[, g$sample, drop = FALSE])
    n <- rowSums(m)
    tibble(block_size = ncol(m),
           all_missing = sum(n == ncol(m)),
           none_missing = sum(n == 0),
           partly_missing = sum(n > 0 & n < ncol(m)))
  }) %>% ungroup() %>% dt_table()
```

Missingness falls mostly in whole plex-run blocks, so the practical unit of
missingness is the plex, not the sample. Imputing a whole-block value would mean
inventing a plex of data, so no imputation is done. If a future dataset instead
showed mostly scattered, within-block missingness, imputation would become
defensible and the method should be chosen from the mechanism (left-censored
MNAR methods such as MinProb/QRILC for below-detection missingness; a neighbour
method such as KNN for missing-at-random).

#### Principal component analysis

```{r pca}
# AIM: compress the proteins to a few summary axes and plot samples on the first
# two, coloured by subtype and by plex, to see whether biology or plex drives the
# variation.

pca_fit <- prot.aml.merge.filtered %>% t() %>% pcaMethods::pca(nPcs = 10)
pca_scores <- pcaMethods::scores(pca_fit) %>% as_tibble(rownames = "label_id") %>%
  dplyr::left_join(meta.aml, by = "label_id")
var_explained <- round(100 * pca_fit@R2, 1)

p <- pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, col = subtype)) +
  geom_point(size = 2.5) +
  labs(x = paste0("PC1, ", var_explained[1], "%"),
       y = paste0("PC2, ", var_explained[2], "%")) +
  cowplot::theme_minimal_grid() + cowplot::panel_border()
ggExtra::ggMarginal(p, groupFill = TRUE, groupColour = TRUE)

pca_plex_plot <- pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, col = factor(plex))) +
  geom_point(size = 2.5) + scale_color_viridis_d(name = "Plex") +
  labs(x = paste0("PC1, ", var_explained[1], "%"),
       y = paste0("PC2, ", var_explained[2], "%")) +
  cowplot::theme_minimal_grid() + cowplot::panel_border()
save_ggplot(pca_plex_plot, "02_pca_by_plex.png", w = 7, h = 5)
pca_plex_plot
```

#### Plex and subtype confounding

```{r confounding}
# AIM: quantify how tangled subtype and plex are. Cramér's V runs 0 (independent)
# to 1 (perfectly associated).
confounding_table <- table(meta.aml$subtype, meta.aml$plex)
confounding_table

cramers_v <- function(tab) {
  chi <- suppressWarnings(chisq.test(tab)$statistic)
  as.numeric(sqrt(chi / (sum(tab) * (min(dim(tab)) - 1))))
}
```

Cramér's V between subtype and plex is `r round(cramers_v(confounding_table), 2)`.
Because subtypes sit inside single plexes, plex cannot simply be adjusted for and
batch-correction on plex would delete biology. The one exception is plex 7.

<!--
  IF subtype were spread across plexes (a LOW Cramér's V), you could instead
  adjust for plex directly, e.g. model.matrix(~ subtype + plex) in a limma fit,
  or limma::removeBatchEffect(mat, batch = plex) for visualisation only. That is
  NOT valid here because subtype and plex are confounded.
-->

```{r plex7}
# AIM: show plex 7, the only plex holding both controls and disease. With 6
# controls and 3 disease samples, it gives a plex-effect-free comparison used in
# the second report.
meta.aml %>% dplyr::filter(plex == 7) %>%
  dplyr::select(label_id, tube_id, subtype, group) %>% dt_table()
```

## Summary and export

```{r summary-numbers}
n_samples  <- ncol(prot.aml.merge.filtered)
n_proteins <- nrow(prot.aml.merge.filtered)
```

The dataset comprises `r n_samples` samples with `r n_proteins` proteins retained
after the missingness filter. Sample loading was verified and internal reference
scaling was applied using the pooled reference channel. The two injections per
sample agree closely (mean r `r round(mean(replicate_agreement$r), 3)`) and were
averaged. No values were imputed. The dominant limitation is subtype-plex
confounding; plex 7 provides the one plex-effect-free contrast.

```{r export}
# AIM: bundle the corrected matrix, sample metadata and protein annotation into a
# SummarizedExperiment and save it for report 2.
meta.export <- meta.aml %>%
  dplyr::filter(label_id %in% colnames(prot.aml.merge.filtered)) %>%
  column_to_rownames("label_id")

aml.summarizedExperiment <- SummarizedExperiment::SummarizedExperiment(
  assays = list("protein_unimputed" = prot.aml.merge.filtered),
  colData = meta.export[colnames(prot.aml.merge.filtered), ],
  rowData = prot.annotation[match(rownames(prot.aml.merge.filtered),
                                  prot.annotation$uniprotid), ])

saveRDS(aml.summarizedExperiment,
        file.path(params_local$out_dir,
                  paste0(Sys.Date(), "_pedPMSproteomics.summarizedExperiment.RDS")))
```

# Session Info

```{r session-info}
sessionInfo()
```
