---
title: "pedPMS Proteomics 2: Exploration and analysis (workshop)"
author: "Nabil Ahmed"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    toc-title: "Table of contents"
    code-fold: true
    code-summary: "Code"
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    theme: cosmo
execute:
  warning: false
  message: false
editor: source
---

<!--
  WORKSHOP EDITION (conflict-hardened), report 2 of 2. It reads the object report
  1 saved, clusters the samples, finds proteins that mark each cluster, tests
  those for pathway enrichment, and runs a plex-effect-free healthy-versus-disease
  comparison inside plex 7. Run report 1 first so the .RDS exists.

  Every chunk starts with "# AIM:". Read %>% as "then". Namespace clashes between
  the tidyverse, base, matrixStats and the Bioconductor stack (igraph,
  AnnotationDbi, clusterProfiler, MatrixGenerics ...) are resolved once in setup.

  NOTE on the synthetic data: gene symbols are fictional, so they do not map to
  org.Hs.eg.db. The enrichment sections therefore return nothing and print a clear
  note instead of failing. On real data with real symbols they return enriched
  terms. Always check the bitr() mapping rate before trusting an enrichment table.
-->

```{r setup}
# AIM: load packages, resolve namespace conflicts, set the seed, find the object
# report 1 wrote, and set analysis thresholds.

library(tidyverse)
library(matrixStats)
library(cowplot)
library(ggrepel)
library(DT)
library(SummarizedExperiment)
library(limma)                 # linear models for differential/signature testing
library(bluster)               # makeKNNGraph(): nearest-neighbour graph
library(igraph)                # graph tools + community detection
library(ComplexHeatmap)        # signature heatmap
library(circlize)              # colorRamp2(): heatmap colour scale
library(clusterProfiler)       # enrichGO(), bitr(): enrichment + ID conversion
library(org.Hs.eg.db)          # human gene annotation database #note: will not work in this script
library(ReactomePA)            # enrichPathway(): Reactome enrichment (optional) #note: will not work in this script
library(here)                  # here(): locate the project root portably

# --- Namespace conflicts -------------------------------------------------------
# This report attaches more Bioconductor packages than report 1 (igraph,
# AnnotationDbi via org.Hs.eg.db, clusterProfiler, MatrixGenerics ...), so there
# are more clashes: select (AnnotationDbi/clusterProfiler vs dplyr), rename/slice/
# mutate (clusterProfiler vs dplyr), rowSds/rowVars (matrixStats vs MatrixGenerics),
# table (base vs BiocGenerics), and so on. Resolve them once, by policy:
#   1) dplyr wins every conflict it is part of.
#   2) pin the base / matrixStats functions we call by bare name.
# If you add a package later and hit a "conflict" error, add one line:
#   conflicts_prefer(thePackageYouWant::theFunction)
library(conflicted)            # needs conflicted >= 1.1.0 for conflict_prefer_all()
suppressMessages({
  conflict_prefer_all("dplyr", quiet = TRUE)
  conflicts_prefer(
    matrixStats::rowSds, matrixStats::rowVars,      # vs MatrixGenerics
    base::table, base::which.max, base::which.min,  # vs BiocGenerics
    base::rowMeans, base::colMeans,
    base::rowSums,  base::colSums,
    stats::median,  stats::sd
  )
})

set.seed(20240101)
ht_opt$message = FALSE   # silence a harmless ComplexHeatmap layout note

# Same project-root logic as report 1, so this report reads results/ from the
# exact folder report 1 wrote to, on any machine and without editing a path.
base_dir <- here::here()

params_local <- list(
  # find the newest object report 1 wrote, by pattern, so the two reports never
  # drift apart on the date in the file name. local({...}) runs the search once.
  se_rds     = local({
    f <- list.files(file.path(base_dir, "results"),
                    pattern = "_pedPMSproteomics\\.summarizedExperiment\\.RDS$",
                    full.names = TRUE)
    if (length(f) == 0) stop("No RDS found in ", file.path(base_dir, "results"),
                             ". Run report 1 first.", call. = FALSE)
    f[which.max(file.mtime(f))]
  }),
  out_dir    = file.path(base_dir, "results"),
  n_var_prot = 1000,
  knn_k      = 10,
  fdr_cutoff = 0.05,
  lfc_cutoff = 1.0
)

dt_table <- function(x) {
  DT::datatable(x, filter = "top",
                options = list(autoWidth = FALSE, scrollX = TRUE),
                class = "compact hover row-border stripe dt-left cell-border nowrap")
}

# AIM: build the output folder tree (shared with report 1). Tables go under
# results/tables, image files under results/figures.
dirs <- list(
  root    = params_local$out_dir,
  tables  = file.path(params_local$out_dir, "tables"),
  figures = file.path(params_local$out_dir, "figures"))
for (p in dirs) dir.create(p, showWarnings = FALSE, recursive = TRUE)

# save a data frame to results/tables and return it unchanged (for piping to dt_table).
save_table <- function(x, name) {
  readr::write_csv(x, file.path(dirs$tables, name))
  x
}
# save a ggplot to results/figures and return it unchanged (still displays).
save_ggplot <- function(p, name, w = 8, h = 5) {
  ggplot2::ggsave(file.path(dirs$figures, name), p, width = w, height = h,
                  dpi = 150, bg = "white")
  p
}
```

# Introduction

This report explores the normalised pedPMS proteome saved by report 1. It groups
samples by their protein profiles, describes each group, and tests those groups
for enriched pathways. It closes with a targeted, plex-effect-free comparison of
the six healthy controls against the disease samples that share their plex.

```{r load-se}
# AIM: load the object and unpack it into the matrix, sample metadata and protein
# annotation, then build the feature matrix used for clustering.
se <- readRDS(params_local$se_rds)
prot.aml       <- assay(se, "protein_unimputed")
meta.aml       <- as_tibble(colData(se), rownames = "label_id")
prot.annotation <- as_tibble(rowData(se))

# lookup from protein id to gene symbol (first symbol if several).
uniprotid_to_gene <- prot.annotation %>%
  dplyr::mutate(gene = str_split_i(gene, ";", 1)) %>%
  dplyr::select(uniprotid, gene) %>% deframe()

# feature matrix: complete proteins, centred, top 1000 most variable.
prot.complete <- prot.aml[complete.cases(prot.aml), ]
prot.centred  <- prot.complete - rowMeans(prot.complete)
top_var <- prot.centred[order(matrixStats::rowVars(prot.centred), decreasing = TRUE)[
  seq_len(params_local$n_var_prot)], ]
```

`r nrow(prot.complete)` proteins are complete across all `r ncol(prot.aml)`
samples; the `r params_local$n_var_prot` most variable are used for clustering.
(Complete-case filtering drops any protein missing in even one plex block, so the
clustered feature set is biased toward ubiquitously detected proteins; this is the
price of not imputing blockwise-missing data.)

# Clustering

```{r knn-and-community}
# AIM: compress to 10 PCs, connect each sample to its 10 nearest neighbours, and
# detect communities (groups more connected internally than externally).

pca_fit <- prcomp(t(top_var), center = TRUE, scale. = FALSE)
latent  <- pca_fit$x[, 1:10]

knn_graph <- bluster::makeKNNGraph(latent, k = params_local$knn_k)
igraph::V(knn_graph)$name <- meta.aml$label_id

communities <- igraph::cluster_louvain(knn_graph)
meta.aml$community <- factor(igraph::membership(communities))

tibble(measure = c("modularity", "n communities"),
       value = c(round(igraph::modularity(communities), 3), length(communities))) %>%
  dt_table()

# cross the communities against subtype and against plex. The community-vs-plex
# table is the honest check: a community that is essentially one plex may be a
# batch artifact rather than biology.
table(meta.aml$subtype, meta.aml$community)
table(meta.aml$plex, meta.aml$community)
```

```{r cluster-stability}
# AIM: check the clustering is stable by re-running it on random 80% subsets of
# the proteins and measuring how often sample pairs change co-membership (Hamming
# distance). Low values mean a trustworthy partition. NB this tests robustness to
# dropping proteins only, not to the choice of k or the number of PCs.
co_membership <- function(cl) outer(cl, cl, "==") * 1
reference_co <- co_membership(as.integer(igraph::membership(communities)))

hamming_runs <- map_dbl(1:100, function(i) {
  idx <- sample(nrow(top_var), floor(0.8 * nrow(top_var)))
  pc_b <- prcomp(t(top_var[idx, ]), center = TRUE, scale. = FALSE)$x[, 1:10]
  cl_b <- igraph::membership(igraph::cluster_louvain(bluster::makeKNNGraph(pc_b, k = params_local$knn_k)))
  mean(co_membership(as.integer(cl_b)) != reference_co)
})

tibble(hamming = hamming_runs) %>%
  ggplot(aes(x = hamming)) + geom_histogram(bins = 25) +
  geom_vline(xintercept = mean(hamming_runs), linetype = "dashed", col = "blue") +
  cowplot::theme_minimal_grid() + cowplot::panel_border() +
  labs(x = "Hamming distance to full-feature partition", y = "",
       caption = paste0("mean ", round(mean(hamming_runs), 3), " over 100 resamplings"))
```

# Cluster investigation

```{r cluster-assignments}
# AIM: show each sample's cluster next to its subtype and plex.
meta.aml %>%
  dplyr::select(label_id, tube_id, subtype, group, plex, community) %>% dt_table()
```

## Signature proteins per cluster

```{r signature-model}
# AIM: for each cluster, find proteins higher (or lower) in it than in all other
# clusters combined, using limma (the standard group-comparison tool).
#
# NOTE on interpretation: cluster membership was derived from these same samples,
# so the p-values / FDR below are optimistic (double-dipping). Treat the output as
# a ranked hypothesis list, not validated inference.

design <- model.matrix(~ 0 + meta.aml$community)
colnames(design) <- paste0("C", levels(meta.aml$community))
fit <- lmFit(prot.complete, design)

n_per <- table(meta.aml$community)
one_vs_rest <- map_chr(colnames(design), function(cn) {
  others <- setdiff(colnames(design), cn)
  w <- n_per[str_remove(others, "C")] / sum(n_per[str_remove(others, "C")])
  paste0(cn, " - (", paste(sprintf("%.6f*%s", w, others), collapse = " + "), ")")
})
names(one_vs_rest) <- colnames(design)
contrast_matrix <- makeContrasts(contrasts = one_vs_rest, levels = design)
colnames(contrast_matrix) <- names(one_vs_rest)
fit2 <- eBayes(contrasts.fit(fit, contrast_matrix))   # eBayes shares variance across proteins

signature_results <- map_dfr(colnames(contrast_matrix), function(cn) {
  topTable(fit2, coef = cn, number = Inf, sort.by = "none", confint = TRUE) %>%
    as_tibble(rownames = "uniprotid") %>%
    dplyr::mutate(community = cn, gene = uniprotid_to_gene[uniprotid]) %>%
    dplyr::rename(FDR = adj.P.Val, P = P.Value)
})

# save the full results and the significant subset (with P and FDR) to tables.
# The significant table is a HYPOTHESIS list (see the double-dipping note above).
signature_results %>%
  dplyr::select(community, gene, uniprotid, logFC, CI.L, CI.R, P, FDR) %>%
  dplyr::arrange(community, P) %>%
  save_table("signature_proteins_all.csv")

signature_significant <- signature_results %>%
  dplyr::filter(FDR < params_local$fdr_cutoff, abs(logFC) > params_local$lfc_cutoff) %>%
  dplyr::select(community, gene, uniprotid, logFC, CI.L, CI.R, P, FDR) %>%
  dplyr::arrange(community, P)
save_table(signature_significant, "signature_proteins_significant.csv")
```

```{r signature-heatmap, fig.height=9}
# AIM: heatmap of the strongest signature proteins, with subtype and plex on top.
signature_ids <- signature_results %>%
  dplyr::filter(FDR < params_local$fdr_cutoff, abs(logFC) > params_local$lfc_cutoff) %>%
  group_by(community) %>% slice_max(abs(logFC), n = 15) %>%
  pull(uniprotid) %>% unique()

mat <- prot.centred[signature_ids, ]
mat <- (mat - rowMeans(mat)) / matrixStats::rowSds(mat)   # z-score each protein
rownames(mat) <- uniprotid_to_gene[signature_ids]

sig_heatmap <- ComplexHeatmap::Heatmap(
  mat, name = "z-score",
  col = circlize::colorRamp2(c(-2.5, 0, 2.5), c("#2166ac", "white", "#b2182b")),
  column_split = meta.aml$community,
  top_annotation = HeatmapAnnotation(subtype = meta.aml$subtype,
                                     plex = factor(meta.aml$plex)),
  row_names_gp = gpar(fontsize = 6), show_column_names = FALSE)

# ComplexHeatmap is not a ggplot, so save it through a PNG device explicitly.
png(file.path(dirs$figures, "03_signature_heatmap.png"),
    width = 9, height = 9, units = "in", res = 150)
ComplexHeatmap::draw(sig_heatmap)
invisible(dev.off())

sig_heatmap    # display in the report
```

## Over-representation analysis (ORA)

```{r ora-setup}
# AIM: set up ORA. The background universe is the tested proteins (converted to
# gene ids), NOT the whole genome, so abundant proteins are not called enriched
# just for being detectable.
#
# Symbols are converted to Entrez ids with bitr(). On the synthetic data the
# symbols are fictional and map to nothing, so the universe is empty; we detect
# that and let the enrichment sections skip cleanly with a clear note.
universe_symbols <- uniprotid_to_gene[rownames(prot.complete)] %>% na.omit() %>% unique()
universe_entrez <- suppressWarnings(
  clusterProfiler::bitr(universe_symbols, "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID)

if (length(universe_entrez) == 0)
  message("No gene symbols mapped to Entrez IDs; ORA will be empty. ",
          "Expected on the synthetic data (fictional symbols). On real data, ",
          "check what fraction of your symbols bitr() maps before trusting ORA.")

hits_for <- function(cn, direction = c("up", "down")) {
  direction <- match.arg(direction)
  g <- signature_results %>%
    dplyr::filter(community == cn, FDR < params_local$fdr_cutoff,
                  if (direction == "up") logFC > params_local$lfc_cutoff
                  else logFC < -params_local$lfc_cutoff) %>%
    pull(gene) %>% na.omit() %>% unique()
  if (length(g) < 10) return(NULL)
  ids <- suppressWarnings(
    clusterProfiler::bitr(g, "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID)
  if (length(ids) < 10) return(NULL)   # too few mapped to test meaningfully
  ids
}
```

```{r ora-go}
# AIM: GO biological-process enrichment for each cluster's up-proteins. Skipped
# cleanly when the universe is empty (fictional symbols) or too few genes map.
go_results <- if (length(universe_entrez) == 0) list() else
  map(set_names(colnames(contrast_matrix)), function(cn) {
    h <- hits_for(cn, "up"); if (is.null(h)) return(NULL)
    clusterProfiler::enrichGO(h, OrgDb = org.Hs.eg.db, ont = "BP",
                              universe = universe_entrez, pAdjustMethod = "BH",
                              qvalueCutoff = 0.1, readable = TRUE)
  }) %>% compact()

if (length(go_results) == 0) {
  cat("No GO enrichment to show. With the synthetic data's fictional gene ",
      "symbols, none map to Entrez IDs, so the enrichment universe is empty. ",
      "On real data with real symbols this section returns enriched terms per ",
      "community.\n", sep = "")
} else {
  map_dfr(names(go_results), ~ as_tibble(go_results[[.x]]) %>% dplyr::mutate(community = .x)) %>%
    dplyr::select(community, Description, GeneRatio, BgRatio, p.adjust, qvalue, geneID) %>%
    save_table("ora_GO_by_community.csv") %>%
    dt_table()
}
```

# NEW: plex-7 healthy-versus-disease, free of plex effect

This is the analysis the extra controls make possible. Plex 7 holds all six
healthy controls and three disease samples, so a comparison confined to plex 7
carries no plex effect at all. Any protein that differs here differs for
biological reasons, not because of the batch. It is a small comparison (nine
samples), so treat it as a focused sensitivity check whose top hits are
candidates for validation, not a definitive disease signature. Note also that
three of the six controls are synthetic and were modelled on the real ones with
smaller sample-to-sample spread, so the number of significant proteins here is
sensitive to their inclusion (see the companion exercise).

```{r plex7-differential}
# AIM: inside plex 7 only, test which proteins differ between the 6 controls and
# the 3 disease samples. Because every sample here is in the same plex, the result
# is free of the plex/batch effect that confounds the whole-cohort comparison.

# pick the plex-7 samples and split them into control vs disease.
p7 <- meta.aml %>% dplyr::filter(plex == 7)
p7_mat <- prot.aml[, p7$label_id]
# keep proteins measured in every plex-7 sample (a fair within-plex comparison).
p7_mat <- p7_mat[complete.cases(p7_mat), ]

grp <- factor(p7$group, levels = c("control", "disease"))
design7 <- model.matrix(~ grp)          # intercept = control; 2nd column = disease effect

# limma with eBayes: stabilises per-protein variance, important with few samples.
fit7 <- eBayes(lmFit(p7_mat, design7))
res7 <- topTable(fit7, coef = 2, number = Inf, confint = TRUE) %>%
  as_tibble(rownames = "uniprotid") %>%
  dplyr::mutate(gene = uniprotid_to_gene[uniprotid]) %>%
  dplyr::rename(FDR = adj.P.Val, P = P.Value) %>%
  dplyr::arrange(P)

cat("plex-7 samples: ", nrow(p7), " (",
    sum(p7$group == "control"), " control, ",
    sum(p7$group == "disease"), " disease); proteins tested: ",
    nrow(p7_mat), "\n", sep = "")

# save the full plex-7 result (every protein, with P and FDR) and the significant
# subset, sorted by P, to results/tables.
res7 %>%
  dplyr::select(gene, uniprotid, logFC, CI.L, CI.R, P, FDR) %>%
  save_table("plex7_healthy_vs_disease_all.csv")

res7 %>%
  dplyr::filter(FDR < params_local$fdr_cutoff, abs(logFC) > params_local$lfc_cutoff) %>%
  dplyr::select(gene, uniprotid, logFC, CI.L, CI.R, P, FDR) %>%
  save_table("plex7_healthy_vs_disease_significant.csv")

# show the top 40 by p-value in the report.
res7 %>% dplyr::select(gene, uniprotid, logFC, CI.L, CI.R, P, FDR) %>%
  dplyr::slice(1:40) %>% dt_table()
```

```{r plex7-volcano, fig.height=5}
# AIM: volcano plot of the plex-7 comparison. x = size of the difference (logFC),
# y = evidence against no-difference (-log10 P). Points top-left/right are the
# strongest candidates.
volcano7 <- res7 %>%
  dplyr::mutate(sig = FDR < params_local$fdr_cutoff & abs(logFC) > params_local$lfc_cutoff) %>%
  ggplot(aes(x = logFC, y = -log10(P), col = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                     name = "passes cutoff") +
  ggrepel::geom_text_repel(
    data = ~ dplyr::slice(dplyr::arrange(.x, P), 1:15),
    aes(label = gene), size = 3, max.overlaps = 20, show.legend = FALSE) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", col = "grey50") +
  cowplot::theme_minimal_grid() + cowplot::panel_border() +
  labs(x = "log2 fold change (disease vs control)", y = "-log10 P",
       title = "Plex-7 healthy vs disease (no plex effect)")
save_ggplot(volcano7, "04_plex7_volcano.png", w = 7, h = 5)
volcano7
```

```{r plex7-enrichment}
# AIM: pathway enrichment on the plex-7 up-in-disease proteins, same tested-
# proteome background as above. Guarded on the ID mapping, not just the hit count,
# so it skips cleanly when the fictional symbols do not map.
p7_up <- res7 %>%
  dplyr::filter(FDR < params_local$fdr_cutoff, logFC > params_local$lfc_cutoff) %>%
  pull(gene) %>% na.omit() %>% unique()

ids <- if (length(p7_up)) suppressWarnings(
  clusterProfiler::bitr(p7_up, "SYMBOL", "ENTREZID", org.Hs.eg.db)$ENTREZID) else character(0)

if (length(ids) >= 10 && length(universe_entrez) > 0) {
  ego7 <- clusterProfiler::enrichGO(ids, OrgDb = org.Hs.eg.db, ont = "BP",
                                    universe = universe_entrez, pAdjustMethod = "BH",
                                    qvalueCutoff = 0.1, readable = TRUE)
  as_tibble(ego7) %>%
    dplyr::select(Description, GeneRatio, BgRatio, p.adjust, qvalue, geneID) %>%
    save_table("plex7_enrichment_GO.csv") %>%
    dt_table()
} else {
  cat("Enrichment skipped:", length(ids), "of", length(p7_up),
      "significant up-in-disease genes mapped to Entrez IDs",
      "(fictional symbols do not map). Expected on the synthetic data.\n")
}
```

# Summary and outlook

```{r summary-numbers}
n_communities <- length(communities)
n_p7_sig <- sum(res7$FDR < params_local$fdr_cutoff &
                abs(res7$logFC) > params_local$lfc_cutoff)
```

Louvain clustering on the `r params_local$n_var_prot` most variable proteins gave
`r n_communities` communities across `r ncol(prot.aml)` samples, with mean Hamming
distance `r round(mean(hamming_runs), 3)` over 100 resamplings. The plex-7
comparison, which is free of plex effect, returned `r n_p7_sig` proteins passing
the fold-change and FDR cutoffs. That count depends on the three synthetic,
low-variance controls; with the three real controls only it drops to near zero, so
it should be read as a candidate list, not a result.

What this report supports and what it does not:

The cluster structure and its signature proteins are exploratory. Membership was
derived from the same data used to test it, so those proteins are a ranked
hypothesis list, not a validated signature. Where a cluster lines up with a single
plex, its signature cannot be attributed to biology from this dataset alone; the
plex annotation on the heatmap is the check for that.

The plex-7 comparison is the one contrast here that is structurally free of plex
effect, which is exactly why the extra controls were added to that plex. It is
small and leans on synthetic controls, so its value is in nominating candidates
for an independent experiment, not in settling anything by itself.

# Session Info

```{r session-info}
sessionInfo()
```
